---
title: "04-DFbetaAnalysis2"
author: "Kevin O'Brien"
date: "Sunday, August 30, 2015"
output: html_document
---

```{r}
#using package NLME
library(nlme)
library(lme4)
source("data/BloodData.R")
```

```{r}
BloodA <- data.frame(Blood[,7:9])
BloodB <- data.frame(Blood[,7:9])
BloodC <- data.frame(Blood[,7:9])
BloodD <- data.frame(Blood[,7:9])
colnames(BloodA) <- c("A1","A2","A3")
colnames(BloodB) <- c("B1","B2","B3")
colnames(BloodC) <- c("C1","C2","C3")
colnames(BloodD) <- c("D1","D2","D3")
```
```{r}
BloodA <- BloodA-15
set.seed(1234); ChangeB <- round(matrix(rnorm(255,15,2),nrow=85),3)
BloodB <- BloodB - ChangeB
set.seed(2345); ChangeC <- round(matrix(rnorm(255,15,5),nrow=85),3)
BloodC <- BloodC - ChangeC
set.seed(3456); ChangeD <- round(matrix(rnorm(255,15,10),nrow=85),3)
BloodD <- BloodD - ChangeD

Blood1 <- data.frame(Blood[,1:3],BloodA,BloodB)
Blood2 <- data.frame(Blood[,1:3],BloodC,BloodD)

Blood1<-as.matrix(Blood1);head(Blood1)
Blood2<-as.matrix(Blood2);head(Blood2)
```
### Working on the Blood Data

```{r}
blood = groupedData( BP ~ method | subject ,
  data = data.frame( BP = c(Blood), 
    subject = rep(seq(nrow(Blood)), ncol(Blood)),
    method = rep(c("J","R","S"), rep(nrow(Blood)*3, 3)),
    repl = rep(rep(c(1:3), rep(nrow(Blood), 3)), 3) ),
  labels = list(BP = "Systolic Blood Pressure", method = "Measurement Device"),
	)
```

```{r}
head(Blood)
head(Blood2)
```

```{r}

blood1 = groupedData( BP ~ method | subject ,
  data = data.frame( BP = c(Blood1), 
    subject = rep(seq(nrow(Blood1)), ncol(Blood1)),
    method = rep(c("J","A","B"), rep(nrow(Blood1)*3, 3)),
    repl = rep(rep(c(1:3), rep(nrow(Blood1), 3)), 3) ),
  labels = list(BP = "Systolic Blood1 Pressure", method = "Measurement Device"),
    )
blood2 = groupedData( BP ~ method | subject ,
  data = data.frame( BP = c(Blood2), 
    subject = rep(seq(nrow(Blood2)), ncol(Blood2)),
    method = rep(c("J","C","D"), rep(nrow(Blood2)*3, 3)),
    repl = rep(rep(c(1:3), rep(nrow(Blood2), 3)), 3) ),
  labels = list(BP = "Systolic Blood2 Pressure", method = "Measurement Device"),
    )
```

```{r}
roy.LME4.jsr = lmer(BP ~ method-1 + (1 |subject ), data = blood,REML = FALSE)
roy.LME4.jab = lmer(BP ~ method-1 + (1 |subject ), data = blood1,REML = FALSE)
roy.LME4.jcd = lmer(BP ~ method-1 + (1 |subject ), data = blood2,REML = FALSE)
```
```{r}
summary(roy.LME4.jab)
```
```{r}
summary(roy.LME4.jcd)
```
```{r}
library(influence.ME)
library(dplyr)
roy.LME4.inf  <- influence(roy.LME4.jab, "subject")
CDs<- cooks.distance(roy.LME4.inf)
case<- rownames(cooks.distance(roy.LME4.inf))
CDdf <- data.frame(case,CDs)
CDdf<- arrange(CDdf,desc(CDs))
head(CDdf)
dfbeta <- dfbetas(roy.LME4.inf)
case <- rownames(dfbetas(roy.LME4.inf))
DFBetadf <- data.frame(case,dfbeta)
Infldf <- inner_join(CDdf,DFBetadf)
Infldf<- arrange(Infldf,desc(CDs))
rownames(Infldf) <- Infldf$case
head(Infldf)
```
```{r}
par(mfrow=c(2,2))
plot(Infldf$methodJ,Infldf$methodA,col="red",pch=16,xlab=c("method J"),ylab=c("method A"))
abline(c(0,1))

plot(Infldf$methodJ,Infldf$methodB,col="red",pch=16,xlab=c("method J"),ylab=c("method B"))
abline(c(0,1))

plot(Infldf$methodA,Infldf$methodB,col="red",pch=16,xlab=c("method A"),ylab=c("method B"))
abline(c(0,1))

par(mfrow=c(1,1))

```


```{r}
library(influence.ME)
library(dplyr)
roy.LME4.inf  <- influence(roy.LME4.jcd, "subject")
CDs<- cooks.distance(roy.LME4.inf)
case<- rownames(cooks.distance(roy.LME4.inf))
CDdf <- data.frame(case,CDs)
CDdf<- arrange(CDdf,desc(CDs))
head(CDdf)
dfbeta <- dfbetas(roy.LME4.inf)
case <- rownames(dfbetas(roy.LME4.inf))
DFBetadf <- data.frame(case,dfbeta)
Infldf <- inner_join(CDdf,DFBetadf)
Infldf<- arrange(Infldf,desc(CDs))
rownames(Infldf) <- Infldf$case
head(Infldf)
```
```{r}
par(mfrow=c(2,2))
plot(Infldf$methodJ,Infldf$methodC,col="red",pch=16,xlab=c("method J"),ylab=c("method C"))
abline(c(0,1))

plot(Infldf$methodJ,Infldf$methodD,col="red",pch=16,xlab=c("method C"),ylab=c("method D"))
abline(c(0,1))

plot(Infldf$methodC,Infldf$methodD,col="red",pch=16,xlab=c("method C"),ylab=c("method D"))
abline(c(0,1))

par(mfrow=c(1,1))

```

```{r}
# Roy's MCS Procedures.

library(nlme)
```

```{r}

dat.ja = subset(blood1, subset = method != "B")

dim(dat.ja)
#Roy's Candidate Models
```

```{r}
dat.roy1 = lme(BP ~ method-1, data = dat.ja,
random = list(subject=pdSymm(~ method-1)), weights=varIdent(form=~1|method),
correlation = corSymm(form=~1 | subject/repl), method="ML")
```
```{r}
dat.roy2 = lme(BP ~ method-1, data = dat.ja,
random = list(subject=pdCompSymm(~ method-1)), 
correlation = corSymm(form=~1 | subject/repl), method="ML")
```
```{r}
dat.roy3 = lme(BP ~ method-1, data = dat.ja,
random = list(subject=pdSymm(~ method-1)),weights=varIdent(form=~1|method), 
correlation = corCompSymm(form=~1 | subject/repl), method="ML")
```
```{r}
dat.roy4 = lme(BP ~ method-1, data = dat.ja,
random = list(subject=pdCompSymm(~ method-1)), 
correlation = corCompSymm(form=~1 | subject/repl), method="ML")
```
```{r}
#ANOVAs
test1 = anova(dat.roy1,dat.roy2) # Between-Subject Variabilities
test2 = anova(dat.roy1,dat.roy3) # Within-Subject Variabilities
test3 = anova(dat.roy1,dat.roy4) # Overall Variabilities
```
### Between-Subject Variabilities
```{r}
test1
```
### Within-Subject Variabilities
```{r}
test2
```
### Overall Variabilities
```{r}
test3
```


```{r}

dat.jb = subset(blood1, subset = method != "A")

dim(dat.jb)

```
### Roy's Candidate Models
```{r}
dat.roy1 = lme(BP ~ method-1, data = dat.jb,
random = list(subject=pdSymm(~ method-1)), weights=varIdent(form=~1|method),
correlation = corSymm(form=~1 | subject/repl), method="ML")
```
```{r}
dat.roy2 = lme(BP ~ method-1, data = dat.jb,
random = list(subject=pdCompSymm(~ method-1)), 
correlation = corSymm(form=~1 | subject/repl), method="ML")
```
```{r}
dat.roy3 = lme(BP ~ method-1, data = dat.jb,
random = list(subject=pdSymm(~ method-1)),weights=varIdent(form=~1|method), 
correlation = corCompSymm(form=~1 | subject/repl), method="ML")
```
```{r}
dat.roy4 = lme(BP ~ method-1, data = dat.jb,
random = list(subject=pdCompSymm(~ method-1)), 
correlation = corCompSymm(form=~1 | subject/repl), method="ML")
```
```{r}
#ANOVAs
test1 = anova(dat.roy1,dat.roy2) # Between-Subject Variabilities
test2 = anova(dat.roy1,dat.roy3) # Within-Subject Variabilities
test3 = anova(dat.roy1,dat.roy4) # Overall Variabilities
```
### Between-Subject Variabilities
```{r}
test1
```
### Within-Subject Variabilities
```{r}
test2
```
### Overall Variabilities
```{r}
test3
```



```{r}

dat.jc = subset(blood2, subset = method != "D")

dim(dat.jc)
#Roy's Candidate Models
```

```{r}
dat.roy1 = lme(BP ~ method-1, data = dat.jc,
random = list(subject=pdSymm(~ method-1)), weights=varIdent(form=~1|method),
correlation = corSymm(form=~1 | subject/repl), method="ML")
```
```{r}
dat.roy2 = lme(BP ~ method-1, data = dat.jc,
random = list(subject=pdCompSymm(~ method-1)), 
correlation = corSymm(form=~1 | subject/repl), method="ML")
```
```{r}
dat.roy3 = lme(BP ~ method-1, data = dat.jc,
random = list(subject=pdSymm(~ method-1)),weights=varIdent(form=~1|method), 
correlation = corCompSymm(form=~1 | subject/repl), method="ML")
```
```{r}
dat.roy4 = lme(BP ~ method-1, data = dat.jc,
random = list(subject=pdCompSymm(~ method-1)), 
correlation = corCompSymm(form=~1 | subject/repl), method="ML")
```
```{r}
#ANOVAs
test1 = anova(dat.roy1,dat.roy2) # Between-Subject Variabilities
test2 = anova(dat.roy1,dat.roy3) # Within-Subject Variabilities
test3 = anova(dat.roy1,dat.roy4) # Overall Variabilities
```
### Between-Subject Variabilities
```{r}
test1
```
### Within-Subject Variabilities
```{r}
test2
```
### Overall Variabilities
```{r}
test3
```


```{r}

dat.jd = subset(blood2, subset = method != "C")

dim(dat.jd)

```
### Roy's Candidate Models
```{r}
dat.roy1 = lme(BP ~ method-1, data = dat.jd,
random = list(subject=pdSymm(~ method-1)), weights=varIdent(form=~1|method),
correlation = corSymm(form=~1 | subject/repl), method="ML")
```
```{r}
dat.roy2 = lme(BP ~ method-1, data = dat.jd,
random = list(subject=pdCompSymm(~ method-1)), 
correlation = corSymm(form=~1 | subject/repl), method="ML")
```
```{r}
dat.roy3 = lme(BP ~ method-1, data = dat.jd,
random = list(subject=pdSymm(~ method-1)),weights=varIdent(form=~1|method), 
correlation = corCompSymm(form=~1 | subject/repl), method="ML")
```
```{r}
dat.roy4 = lme(BP ~ method-1, data = dat.jd,
random = list(subject=pdCompSymm(~ method-1)), 
correlation = corCompSymm(form=~1 | subject/repl), method="ML")
```
```{r}
#ANOVAs
test1 = anova(dat.roy1,dat.roy2) # Between-Subject Variabilities
test2 = anova(dat.roy1,dat.roy3) # Within-Subject Variabilities
test3 = anova(dat.roy1,dat.roy4) # Overall Variabilities
```
### Between-Subject Variabilities
```{r}
test1
```
### Within-Subject Variabilities
```{r}
test2
```
### Overall Variabilities
```{r}
test3
```