\documentclass[12pt, a4paper]{report}
\usepackage{epsfig}
\usepackage{subfigure}
%\usepackage{amscd}
\usepackage{amssymb}
\usepackage{framed}
\usepackage{subfiles}
\usepackage{amsbsy}
\usepackage{graphicx}
\usepackage{amsthm}
%\usepackage[dvips]{graphicx}
\usepackage{natbib}
\usepackage{subfiles}
\bibliographystyle{chicago}
\usepackage{vmargin}
\usepackage{index}
% left top textwidth textheight headheight
% headsep footheight footskip
\setmargins{3.0cm}{2.5cm}{15.5 cm}{22cm}{0.5cm}{0cm}{1cm}{1cm}
\renewcommand{\baselinestretch}{1.5}
\pagenumbering{arabic}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{ill}[theorem]{Example}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{axiom}{Axiom}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{notation}{Notation}
\theoremstyle{remark}
\newtheorem{remark}{Remark}[section]
\newtheorem{example}{Example}[section]
\renewcommand{\thenotation}{}
\renewcommand{\thetable}{\thesection.\arabic{table}}
\renewcommand{\thefigure}{\thesection.\arabic{figure}}
\title{Research notes: linear mixed effects models}
\author{ } \date{ }


\makeindex
\begin{document}
	\author{Kevin O'Brien}
	\title{November 2011 Version A}
	
	
	\addcontentsline{toc}{section}{Bibliography}
	
	
	\tableofcontents \setcounter{tocdepth}{1}
	
\begin{framed} 
\begin{itemize}
	\item \texttt{R} command and \texttt{R} object - Typewriter Font
	\item \texttt{R} Package name - Italics
	\item Selected Acronyms and Proper Nouns - Italics
\end{itemize}
\end{framed}
\newpage


This chapter is broken into two parts. The first part is a review of diagnostics methods for linear models, intended to acquaint the
reader with the subject, and also to provide a basis for material covered in the second part. Particular attention is drawn to graphical methods.

The second part of the chapter looks at diagnostics techniques for LME models, firsly covering the theory, then proceeding to a discussion on 
implementing these using \texttt{R} code.
While a substantial body of work has been developed in this area, ther are still area worth exploring. 
In particular the development of graphical techniques pertinent to LME models should be looked at.
\newpage
%\subfile{1.1-InfluenceIntro.tex}
%\subfile{1.1-ResidualPlots.tex}
%\subfile{1.2-Studentization.tex}
%%\subfile{1.2-CondMargResidual.tex}
%\subfile{1.2-Influence-OLS.tex}
%%\subfile{1.3-Influence1.tex}
%\subfile{1.4-InfluenceAnalysis.tex}
%%\subfile{1.6-Influence-LME.tex}
%\subfile{1.7-MeasureOfInfluence.tex}
%%\subfile{1.8-Schabenberger.tex}
%\subfile{2.1-Zewotir.tex}
%\subfile{2.3-LindMun-LMEdiag.tex}
%\subfile{2.4-CaseDeletionDiagnostics.tex}
%\subfile{2.5-CooksDistance-OLS.tex}
%\subfile{2.7-InfluenceCooksDistance.tex}
%\subfile{2.8-CooksDistance-LME.tex}
%\subfile{2.6-CPJ.tex}
%%\subfile{3.1-LikelihoodDistances.tex}
%\subfile{3.2-Leverage.tex}
%\subfile{3.3-NobreSinger-LMELeverage.tex}
%\subfile{5.2-HaslettDillane.tex}
%\subfile{5.3-HaslettHayes.tex}
%\subfile{4.1-Turkan-InfluenceAnalysis.tex}
%\subfile{6.3-Influence.ME-Rpackage.tex}

%===================================================================%
\newpage
\begin{itemize}
\item \textit{
	The previous Section (Section 4) is a literary review of residual diagnostics and influence procedures
	for Linear Mixed Effects Models, drawing heavily on Schabenberger and Zewotir.}
	
\item \textit{	Section 4 begins with an introduction to key topics in residual diagnostics, such as influence, leverage, outliers
	and Cook's distance. Other concepts such as DFFITS and DFBETAs will be introduced briefly, mostly to explain why the are not particularly useful for
	the Method Comparison context, and therefore are not elaborated upon.}
	
\item \textit{	In brief, Variable Selection is not applicable to Method Comparison Studies, in the 
	commonly used used context. 
	Testing a rather simplisticy specificied model against one with more random effects terms is tractable, but this research question is of secondary importance.}
\end{itemize}

%=============================================== %
\newpage
\subsection*{Appendix to Section 4}

As an appendix to section 4, an appraisal of the current state of development (or lack thereof) for current implemenations for LME models, particularly for
\texttt{nlme} and \texttt{lme4} fitted models.

Crucially, a review of internet resources indicates that almost all of the progress in this regard has been done for \texttt{lme4} fitted models, specifically the \textit{Influence.ME} \texttt{R} package. (Nieuwenhuis et 2012)

Conversely there is very little for \texttt{nlme} models. To delve into this mor, one would immediately investigate the current development workflow for both packages.

%======================%
% Douglas Bates

As an aside, Douglas Bates was arguably the most prominent \texttt{R} developer working in the LME area. 
However Bates has now prioritised the development of LME models in another computing environment , i.e Julia. 
% The current version of this is XXXX

%======================%
% nlme
\subsubsection*{The \texttt{nlme} package}

With regards to \texttt{nlme}, the torch has been passed to Galecki Galecki \& Burzykowski (UMich. and Hasselt respecitely).  Galecki \& Burzykowski published \textit{Linear Mixed Effects Models using \texttt{R}}. 
Also, the accompanying \texttt{R} package, nlmeU package is under current development, with a version being released XXXX.


%======================%
% lme4 and influence.ME
\subsubsection*{The \texttt{lme4} package}

The \texttt{lme4} package is also under active development, under the leadership of Ben Bolker (McMaster University). According to CRAN, the LME4 package, fits linear and generalized linear mixed-effects models

\begin{quote}
	The models and their components are represented using S4 classes and methods. The core computational algorithms are implemented using the Eigen C++ library for numerical linear algebra and RcppEigen "glue".
	(CRAN)
\end{quote}

%=====================%
% Important Consideration for MCS

The key issue is that \texttt{nlme} allows for the particular specification of Roy's Model, speciifically direct spefiication of the VC matrices for within subject and between subject residuals.
The \texttt{lme4} package does not allow for this.
To advance the ideas that eminate from Roys' paper, one is required to use the \texttt{nlme} context. However, to take advantage of the infrastructure already provided for \texttt{lme4} models, one may change the research question away from that of Roy's paper. 
To this end, an exploration of what textit{influence.ME} can accomplished is merited.





%====================================================================%
\newpage
%\subfile{LikelihoodDistances.tex}



\newpage
\bibliography{DB-txfrbib}
\end{document}


